Towards Comparable Active Learning

### 2023-04-17 doc/al_benchmark_paper.tex
dropout classifier

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Overview}
Table \ref{tab:benchmark_comparison} shows a feature comparison between our proposed benchmark and several existing benchmarks in the literature, as well as methodological AL papers with an extensive experiments section. \\
{\color{red} TODO} We include in this table any methodological paper that experiments on at least two domains.
\begin{table}[h]
	\centering
	\begin{tabular}{l | c c c c c}
		Paper & \# Datasets & Domains & Scenarios & Oracle & RL Alg. \\
		\hline
		Beck et al. \cite{beck2021effective} & 4 & 1 & 3 & - & - \\
		Hu et al. \cite{hu2021towards} & 5 & 2 & 1 & - & - \\
		Li et al. \cite{li2022empirical} & 5 & 1 & 1 & - & - \\
		\textbf{Ours} & 5 & 2 & 2 & \checkmark & -
	\end{tabular}
	\caption{Comparison of our benchmark with the existing literature}
	\label{tab:benchmark_comparison}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Problem Description / Delineation}
\textbf{Basic Classification}\\
We assume a dataset $\mathcal{D} := (x_i, y_i) \hspace{0.5mm}; \hspace{1mm} i := 1 \ldots N$ consisting of instances $x_i \in \mathbb{R}^M$ and corresponding $y_i \in \mathbb{R}^C$.
For evaluation purposes we assume a held-out test set $\mathcal{D}^{\text{test}}$ with the same characteristics.
We consider classification problems with one-hot encoded classes, hence $C$ models the number of classes.
To perform classification, a model $\hat y_\theta : \mathbb{R}^M \rightarrow \mathbb{R}^C$ is used. To fit the model, it is parameterized by $\theta$ and subjected to loss $\ell: \mathbb{R}^C \times \mathbb{R}^C \rightarrow \mathbb{R}$. For this work, categorical cross-entropy (CE) is used.
For evaluating classification performance, we use accuracy on the test set $\operatorname{Acc}(\mathcal{D}^{\text{test}}, \hat{y}_\theta)$. \\ [1mm]
%
\textbf{Pool-based AL with single instances (non-batch setting)}\\
To construct the active learning setting, we suppress the labels $y_i$ of $\mathcal{D}$ to form the unlabeled pool $\mathcal{U} := u_i \hspace{0.5mm}; \hspace{1mm} i := 1 \ldots N$ and form and initial labeled pool $\mathcal{L}$ by uniformly sampling $k$ number of instances per class from $\mathcal{U}$ and recovering their label. 
The result of this so-called "seeding" process is $\mathcal{L} := (u_i, y_i) \hspace{0.5mm}; \hspace{1mm} i := 1 \ldots k*C$. \\
Active learning is defined as sequentially removing single instances $u^{(i)} \in \mathcal{U}^{(i)}\hspace{1mm}; \hspace{2mm} \mathcal{U}^{(i+1)} := \mathcal{U}^{(i)} \setminus´ \{u^{(i)}\}$, recovering their label $y^{(i)}$ and adding them to the labeled pool $\mathcal{L}^{(i+1)} := \mathcal{L}^{(i)} \cup (u^{(i)}, y^{(i)})$ until a fixed budget $B$ is exhausted $i := 1 \ldots B$.
After each added instance the classification model is retrained according to section \ref{sec:training_the_classifier} and its performance is measured on the held-out test set $\mathcal{D}^{\text{test}}$.
The quality of an active learning algorithm is evaluated by an "anytime" protocol that incorporates classification performance at every iteration, not just the final performance after the budget is exhausted.
We employ the normalized area under the accuracy curve (AUC):
\begin{equation}\label{eq:auc}
	\operatorname{auc}(\mathcal{U}, \mathcal{L}, \hat y, B) := \frac{1}{B} \sum_{i=1}^{B} \operatorname{Acc}(y_{test}, \hat y_i(x_{test}))
\end{equation}
Where $\hat y_i$ is the retrained classification model after the i-\textit{th} instance was selected. \\ [1mm]
%
\textbf{Framing AL as RL}\\
We define the active learning process as an adapted reinforcement learning loop $(S, A, \tau, \Omega, \omega)$ where an environment iteratively will expose a state $s \in S$ to an agent $\Omega$, which will choose actions $a \in A$.
For each iteration $i$ the environment samples a subset of size $\tau$ of unlabeled instances $u^{(i)} \underset{\tau}{\sim} \mathcal{U}^{(i)}$, constructs the state $s^{(i)} := \omega(u^{(i)})$ and presents it to the agent to select an action $a^{(i)} := \Omega(s^{(i)})$.
The action $a^{(i)}$ is the index of the selected instance in $u^{(i)}$ out of all possible indices $A := [1 \ldots \tau]$.
This process is repeated $B$ times $i := [1 \ldots B]$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Reproducibility}\label{sec:reproducibility}
A big focus in this work is to provide an experimental setup that is fully reproducible independent from the dataset, classification model or AL algorithm used.
In our opinion, an evaluation on one dataset with a given seed should always be done on the same sequence of batches $x_0, \ldots, x_B$.
Even though different AL algorithms will pick different samples from these batches, making them unavailable for sampling in later batches, the theoretical decision tree for every possible choice in every iteration $i$ should stay the same.
This behavior is not possible with the default choice of setting a global seed at the start of the experiment, since a single additional random draw from the random number generator completely changes the decision tree for the batches.
This additional random number might be drawn during the initialization of the classification model, or the AL algorithm, or even during every AL iteration if $\Omega$ is stochastic.
This problem also applies to the initialization of the classification models $\theta$, the initialization and querying of $Omega$ and the drawn mini batches during the training of $\theta$.
The desired control over all these processes can be archived by assigning a separate seeded random number generator to all these processes.
To the best of our knowledge, we are the first work that discusses this issue and proposes a solution for it.
We hypothesize that the insufficient setup with global seeds contributes to the on-going problem of inconsistent results of AL algorithms in different papers.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Overview}
Table \ref{tab:benchmark_comparison} shows a feature comparison between our proposed benchmark and several existing benchmarks in the literature, as well as methodological AL papers with an extensive experiments section. \\
{\color{red} TODO} We include in this table any methodological paper that experiments on at least two domains. \\
{\color{red} TODO} Define AL scenarios (really hard)
\begin{table}[h]
	\centering
	\begin{tabular}{l | c c c c c}
		Paper & \# Datasets & Domains & Scenarios & Oracle & RL Alg. \\
		\hline
		Beck et al. \cite{beck2021effective} & 4 & 1 & 3 & - & - \\
		Hu et al. \cite{hu2021towards} & 5 & 2 & 1 & - & - \\
		Li et al. \cite{li2022empirical} & 5 & 1 & 1 & - & - \\
		\textbf{Ours} & 5 & 2 & 2 & \checkmark & -
	\end{tabular}
	\caption{Comparison of our benchmark with the existing literature}
	\label{tab:benchmark_comparison}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Problem Description / Delineation}
\textbf{Lars' Version} \\
Given
\begin{itemize}
	\item a number $B\in\mathcal{N}$ (called budget),
	\item two spaces $\mathcal{X}$ and $\mathcal{Y}$,  {\tiny e.g., $\mathcal{X}:=\mathcal{R}^M, \mathcal{Y}:=\mathcal{R}^T$},
	\item a sample $\mathcal{D}_1,\ldots,\mathcal{D}_N \subseteq (\mathcal{X}\times \mathcal{Y})^*$ of
	sequences of pairs $(x,y)$  from an unknown distribution $p$
	(called datasets),
	%  {\tiny with $|\mathcal{D}_n|\geq B$ for all $n\in 1{:}N$,}
	{\tiny with $p(\mathcal{D})=0$ for $|\mathcal{D}|<B$,}
	\item a function $\ell:\mathcal{Y}\times \mathcal{Y}\rightarrow \mathcal{R}$ (called loss), and
	\item a function $A:  (\mathcal{X} \times \mathcal{Y})^* \times \mathcal{X}^* \rightarrow \mathcal{Y}^{\mathcal{X}}$
	(called learning algorithm),
\end{itemize}
find a function
\begin{align*}
	a: (\mathcal{X}\times \mathcal{Y})^* \times \mathcal{X}^* \rightarrow \{0,1\}^*
	\quad\quad \text{\tiny (equivariant in the second argument)}
\end{align*}
% which is equivariant in the second argument, 
called acquisition function,
s.t. the expected loss of a model learned on
all predictors 
plus $B$ sequentially acquired
targets is minimal:
\begin{align*}
	\mathbb{E}_{\mathcal{D}\text{train},\mathcal{D}\text{test}\sim p}   &
	% \frac{1}{|\mathcal{D}\text{test}|} \sum_{(x,y)\in\mathcal{D}\text{test}}
	% \operatorname{avg} \{
	%  \ell(y, \hat y(x)) \mid (x,y)\in\mathcal{D}\text{test} \}
	\operatorname{avg}\limits_{(x,y)\in\mathcal{D}\text{test}}
	\ell(y, \hat y(x)) 
	\\
	\text{with }
	\hat y:= & A( (\mathcal{D}\text{train}_{n_1},\ldots,\mathcal{D}\text{train}_{n_B}), \mathcal{D}\text{train}|_{\mathcal{X}})
	\\ 
	n_b := & \text{index}( a( (\mathcal{D}\text{train}_{n_1},\ldots,\mathcal{D}\text{train}_{n_{b-1}}), \mathcal{D}\text{train}|_{\mathcal{X}}) ),
	\quad b\in 1{:}B
\end{align*}
\textbf{Notes:}
\begin{itemize}
	\item We would need to switch from lowest expected loss to highest AUC
	\item 
\end{itemize}
\vspace{5mm}
\textbf{Thorben's Version}\\
\textbf{Basic Classification}\\
We assume a dataset $\mathcal{D} := (x_i, y_i) \hspace{0.5mm}; \hspace{1mm} i := 1 \ldots N$ consisting of instances $x_i \in \mathbb{R}^M$ and corresponding $y_i \in \mathbb{R}^C$.
For evaluation purposes we assume a held-out test set $\mathcal{D}^{\text{test}}$ with the same characteristics.
We consider classification problems with one-hot encoded classes, hence $C$ models the number of classes.
To perform classification, a model $\hat y_\theta : \mathbb{R}^M \rightarrow \mathbb{R}^C$ is used. To fit the model, it is parameterized by $\theta$ and subjected to loss $\ell: \mathbb{R}^C \times \mathbb{R}^C \rightarrow \mathbb{R}$. For this work, categorical cross-entropy (CE) is used.
For evaluating classification performance, we use accuracy on the test set $\operatorname{Acc}(\mathcal{D}^{\text{test}}, \hat{y}_\theta)$. \\ [1mm]
%
\textbf{Pool-based AL with single instances (non-batch setting)}\\
To construct the active learning setting, we suppress the labels $y_i$ of $\mathcal{D}$ to form the unlabeled pool $\mathcal{U} := u_i \hspace{0.5mm}; \hspace{1mm} i := 1 \ldots N$ and form and initial labeled pool $\mathcal{L}$ by uniformly sampling $k$ number of instances per class from $\mathcal{U}$ and recovering their label. 
The result of this so-called "seeding" process is $\mathcal{L} := (u_i, y_i) \hspace{0.5mm}; \hspace{1mm} i := 1 \ldots k*C$. \\
Active learning is defined as sequentially removing single instances $u^{(i)} \in \mathcal{U}^{(i)}\hspace{1mm}; \hspace{2mm} \mathcal{U}^{(i+1)} := \mathcal{U}^{(i)} \setminus´ \{u^{(i)}\}$, recovering their label $y^{(i)}$ and adding them to the labeled pool $\mathcal{L}^{(i+1)} := \mathcal{L}^{(i)} \cup (u^{(i)}, y^{(i)})$ until a fixed budget $B$ is exhausted $i := 1 \ldots B$.
After each added instance the classification model is retrained according to section \ref{sec:training_the_classifier} and its performance is measured on the held-out test set $\mathcal{D}^{\text{test}}$.
The quality of an active learning algorithm is evaluated by an "anytime" protocol that incorporates classification performance at every iteration, not just the final performance after the budget is exhausted.
We employ the normalized area under the accuracy curve (AUC):
\begin{equation}\label{eq:auc}
	\operatorname{auc}(\mathcal{U}, \mathcal{L}, \hat y, B) := \frac{1}{B} \sum_{i=1}^{B} \operatorname{Acc}(y_{test}, \hat y_i(x_{test}))
\end{equation}
Where $\hat y_i$ is the retrained classification model after the i-\textit{th} instance was selected. \\ [1mm]
%
\textbf{Framing AL as RL}\\
We define the active learning process as an adapted reinforcement learning loop $(S, A, \tau, \Omega, \omega)$ where an environment iteratively will expose a state $s \in S$ to an agent $\Omega$, which will choose actions $a \in A$.
For each iteration $i$ the environment samples a subset of size $\tau$ of unlabeled instances $u^{(i)} \underset{\tau}{\sim} \mathcal{U}^{(i)}$, constructs the state $s^{(i)} := \omega(u^{(i)})$ and presents it to the agent to select an action $a^{(i)} := \Omega(s^{(i)})$.
The action $a^{(i)}$ is the index of the selected instance in $u^{(i)}$ out of all possible indices $A := [1 \ldots \tau]$.
This process is repeated $B$ times $i := [1 \ldots B]$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Reproducibility}\label{sec:reproducibility}
{\color{red} Version: 1.0}\\
A big focus in this work is to provide an experimental setup that is fully reproducible independent from the dataset, classification model or AL algorithm used.
In our opinion, an evaluation on one dataset with a given seed should always be done on the same sequence of batches $x_0, \ldots, x_B$.
Even though different AL algorithms will pick different samples from these batches, making them unavailable for sampling in later batches, the theoretical decision tree for every possible choice in every iteration $i$ should stay the same.
This behavior is not possible with the default choice of setting a global seed at the start of the experiment, since a single additional random draw from the random number generator completely changes the decision tree for the batches.
This additional random number might be drawn during the initialization of the classification model, or the AL algorithm, or even during every AL iteration if $\Omega$ is stochastic.
This problem also applies to the initialization of the classification models $\theta$, the initialization and querying of $Omega$ and the drawn mini batches during the training of $\theta$.
The desired control over all these processes can be archived by assigning a separate seeded random number generator to all these processes.
To the best of our knowledge, we are the first work that discusses this issue and proposes a solution for it.
We hypothesize that the insufficient setup with global seeds contributes to the on-going problem of inconsistent results of AL algorithms in different papers.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Comparibility}\label{sec:comparibility}
{\color{red} Version: braindump}\\
We split the evaluation model that generates accuracy scores and the model that generates features for the AL agent. \\
This allows for dropout / ensemble approaches to invest more compute to generate features \\
The normal evaluation model is available as a fallback of course \\
This way, the comparison is always fair
